---
title: "Week 3 Assignment - more on data tranformation and ggplot"
date: "Due 2/18/2021"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

Required packages for this assignment

```{r, echo=FALSE, messsage = FALSE, warning=FALSE}
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(lubridate))
```

```{r, echo = TRUE}
library(tidyverse)
library(lubridate)
```

# Plotting COVID trends

For this assignment you will be plotting trends in the COVID-19 pandemic across Iowa and other states in the US.

## Load and clean COVID-19 data

Start by loading in daily COVID-19 data. Data on COVID-19 cases are available from a number of sources; however, one good source is the COVID Tracking Project. Daily data for all 50 states can be found at this address: <https://api.covidtracking.com/v1/states/daily.csv>. Load this data by directly linking to the online .csv file (you can put this path directly into the `read_csv()` function. Once read in, your data should look like the following:

```{r}
# covid_data <- read_csv("https://api.covidtracking.com/v1/states/daily.csv")
# save(covid_data,file = "data/covid/covid_tracking.RData")
load(file = "data/covid/covid_tracking.RData")
covid_data
```

Next load in the state population data available on the ICON website called `state_pop_data.RData`.

```{r, include=FALSE}
#load("~/OneDrive - University of Iowa/R_projects/covid19/data/state_pop_data.RData")
#state_pop_data <- state_pop_data %>% rename(state=location)
#save(state_pop_data,file = "data/covid/state_pop_data.RData")
load("data/covid/state_pop_data.RData")
```

For this assignment, we are going to reduce the dataset down to a handful of variables of interest. The dataset contains cumulative values (e.g., total cases/death) and new values each day (e.g., new test or cases). I am going to select the values `date, state, positive, totalTestResults, death, hospitalizedCurrently, inIcuCurrently, onVentilatorCurrently, positiveIncrease, totalTestResultsIncrease and deathIncrease`. I also changed the names of the variables slightly to make them a bit more interpretable (see following output). Next, merge in the state population data (described above). Finally change the date value from the number value to a date using the `ymd()` function in the `lubridate` package.

Here is an example of what the dataset looks like after selecting and changing variable names and merging in the state population:

```{r}
covid_data <- covid_data %>% 
  select(date,state,
         total_cases = positive,
         total_tests = totalTestResults,
         total_deaths = death,
         current_hospitalized = hospitalizedCurrently,
         current_in_icu = inIcuCurrently,
         current_on_ventilator = onVentilatorCurrently,
         new_cases = positiveIncrease,
         new_tests = totalTestResultsIncrease,
         new_deaths = deathIncrease) %>% 
  inner_join(state_pop_data,by = "state") %>% 
  mutate(date=ymd(date))
covid_data
```

**Optional Step:** At this point you may want to save the cleaned data to a file to make it easier to load and return to in the future (without having to constantly query the COVID Tracking Project website).

# Plotting COVID trends in IA

Now look at trends in COVID-19 in the state of Iowa

## Plotting deaths across time in the state of IA

Start by plotting the cumulative number of deaths across time in the state of Iowa. Use the functions `ggtitle()`, `xlab()` and `ylab()` to add a title and meaningful axis labels

```{r, warning=FALSE}
covid_data %>%
  filter(state=="IA") %>%
  ggplot(aes(date, total_deaths)) +
  geom_line() +
  ggtitle("Total COVID-19 Deaths in Iowa") +
  xlab("Date") +
  ylab("Total Deaths")
```

Next try plotting the number of new deaths each day:

```{r, warning=FALSE}
covid_data %>%
  filter(state=="IA") %>%
  ggplot(aes(date, new_deaths)) +
  geom_line() + 
  ggtitle("New Daily Deaths Reported") +
  xlab("Reporting Date") +
  ylab("New Deaths Reported")
```

Notice that the above plot looks a bit noisy, the data is largely impacted by day of the week reporting effects. Two ways to see through the noise are to view the results at a weekly level or to compute a moving average. For example, you could compute a daily average number of deaths each reporting week (i.e., sum up all deaths in the week then divide by 7). Alternatively, you could create a moving average over the previous 7 days (i.e., for each date add up the previous 7 days and divide by 7).

Try plotting both the weekly average and then a moving average. Hints: (1) For the weekly average, you can create a week indicator value using the `week()` or `epiweek()` functions in the `lubridate` package. Make sure you also keep track of the year, now that it is 2021, you can also use the `year()` function for this. (2) To manually compute the moving average you can use the `lag(x,n)` function in the `dplyr` package to return the n-period lag of the variable x (e.g., `lag(var,2)` would return the 2-period lag of the variable `var`), so you could use the lag function 7 times in the `mutate()` function to create a 7-day moving average. (3) As an alternative, you can install the `zoo` package and use the function `rollmean(new_deaths,7, na.pad = T)` to compute a 7 day moving average inside a mutate statement.

Here is what the weekly average plot should look like:

```{r}
covid_data %>%
  filter(state=="IA") %>% 
  group_by(state,year(date),epiweek(date)) %>% 
  mutate(week_avg_deaths=sum(new_deaths)/7) %>% 
  ungroup() %>% 
  ggplot(aes(date, week_avg_deaths)) +
  geom_line() +
  ggtitle("Weekly-Average Number of Daily Deaths Reported") +
  xlab("Reporting Week") +
  ylab("Weekly average")
```

Here is what the moving average plot should look like:

```{r, warning=FALSE}
covid_data %>%
  filter(state=="IA") %>%
  mutate(new_deaths2 = zoo::rollmean(new_deaths,7, na.pad = T)) %>%
  ggplot(aes(date, new_deaths2)) +
  geom_line() +
  ggtitle("Moving Average of Daily Deaths Reported") +
  xlab("Reporting Date") +
  ylab("7-day moving average")
```

## Positivity rate

Next compute the daily positivity rate (percent of tests that are positive) each day in the state and then plot a histogram of the daily positivity rates:

```{r,  warning=FALSE}
covid_data %>% 
  filter(state=="IA") %>%
  mutate(new_positivity=100*new_cases/new_tests) %>% 
  ggplot(aes(new_positivity)) +
  geom_histogram()

```

The above plot may be a bit noisy because of unusual reporting dates (e.g., days with an unusually large/small number of tests or positive cases). We could instead look at the distribution of weekly positivity rates (i.e. positivity across an entire week). Plot a histogram weekly positivity rates:

```{r,  warning=FALSE}
covid_data %>% 
  filter(state=="IA") %>%
  group_by(year(date),epiweek(date)) %>% 
  mutate(week_positivity=100*sum(new_cases)/sum(new_tests)) %>% 
  ggplot(aes(week_positivity)) +
  geom_histogram()
```

Perhaps a better way to visualize the data is using a box plot (`geom_boxplot()`):

```{r,  warning=FALSE}
covid_data %>% 
  filter(state=="IA") %>%
  group_by(year(date),epiweek(date)) %>% 
  mutate(week_positivity=100*sum(new_cases)/sum(new_tests)) %>% 
  ggplot(aes(week_positivity)) +
  geom_boxplot()
```

**Challenge problem:** How could you build this graph?

```{r,  warning=FALSE}
covid_data %>% 
  filter(state=="IA") %>%
  mutate(`Daily Rates`=100*new_cases/new_tests) %>% 
  group_by(year(date),epiweek(date)) %>% 
  mutate(`Weekly Rates`=100*sum(new_cases)/sum(new_tests)) %>% 
  ungroup() %>% 
  select(`Daily Rates`,`Weekly Rates`) %>% 
  gather() %>% 
  ggplot(aes(x=value,y =key)) +
  geom_violin() +
  xlab("Positivity Rates") +
  ylab("") +
  ggtitle("Distribution of positivity rates in Iowa")
```

## Cases and Deaths

Now try plotting total cases and deaths across time in the state of Iowa. Start by trying this with cases in black and deaths in read. Hint: to do this you can use `geom_line()` twice while specifying the y aesthetic in each `geom_line()` function (e.g., `geom_line(aes(y = ))` )

```{r, warning=FALSE}
covid_data %>% 
  filter(state=="IA") %>% 
  ggplot(aes(date)) +
  geom_line(aes(y=total_cases)) +
  geom_line(aes(y=total_deaths),color = "red") +
  ylab("Total Cases or Deaths") +
  xlab("Date") +
  ggtitle("Trends in cases and deaths in Iowa")
```

It's hard to see the trend in deaths so we could center and scale (i.e., subtract mean and divide by SD) the two lines so they are on the same scale. Hint: you can use one `mutate_at()` function to center and scale the data for these two values, and the `scale()` function can be used to do the scaling.

```{r,warning=FALSE}
covid_data %>% 
  filter(state=="IA") %>% 
  mutate_at(vars(c(total_cases,total_deaths)),~scale(.)) %>% 
  ggplot(aes(date)) +
  geom_line(aes(y=total_cases)) +
  geom_line(aes(y=total_deaths),color = "red") +
  ylab("Normalized Value in Total Cases or Deaths") +
  xlab("Date") +
  ggtitle("Trends in cases and deaths in Iowa")
```

As an alternative you could reshape the data so that you have one column that gives you the key for the different values (i.e.. total cases or total deaths) and then a column with the corresponding values. Then you could use something like facets to compare the two different values in separate plots.

```{r, warning=FALSE}
covid_data %>% 
  filter(state=="IA") %>%
  select(date, total_cases,total_deaths) %>% 
  gather(key = key, value = value, -date) %>%
  mutate(key = ifelse(key == "total_cases","Total Cases", "Total Deaths")) %>% 
  ggplot(aes(date,value)) +
  geom_line() +
  facet_wrap(~key,scales = "free_y") +
  ylab("Total Cases or Deaths") +
  xlab("Date") +
  ggtitle("Trends in cases and deaths in Iowa")
```

# Plotting Trends Across States

## Iowa and its neighbors

Now let's compare COVID data in Iowa to neighboring states.

Let's start by repeating the box plot created above to show the distribution of weekly positivity rates across states. Specifically, use multiple box plots that are distinguished by color (note: here we use both the x and y arguments of the boxplot function).

```{r, warning = FALSE}
covid_data %>%
  filter(state %in% c("IA","NE","SD","WI","MN","IL","MO")) %>% 
  group_by(state,year(date),epiweek(date)) %>% 
  mutate(week_positivity=100*sum(new_cases)/sum(new_tests)) %>% 
  ggplot(aes(week_positivity,state, fill = state)) +
  geom_boxplot() +
  ylab("State") +
  xlab("Weekly Positivity") +
  ggtitle("Weekly Positivity Rates Across States")
```

Next return to daily death count line plot from above, and compare across states using colors to distinguish states:

```{r}
covid_data %>%
  filter(state %in% c("IA","NE","SD","WI","MN","IL","MO")) %>% 
  ggplot(aes(date,new_deaths,color = state)) +
  geom_line() +
  ylab("New Daily Deaths") +
  xlab("Date") +
  ggtitle("New daily deaths reported in IA and neighboring states")
```

That was ugly! It's a bit challenging to see a difference in the trends across states because they are overlapping. An alternative approach is to plot a grid of individual plots for each state:

```{r}
covid_data %>%
  filter(state %in% c("IA","NE","SD","WI","MN","IL","MO")) %>% 
  ggplot(aes(date,new_deaths)) +
  geom_line() +
  facet_wrap(~state, scales = "free_y")+
  ylab("New Daily Deaths") +
  xlab("Date") +
  ggtitle("New daily deaths reported in IA and neighboring states")

```

Notice that NE and MN still have some weirdness going on (i.e., negative deaths). Clean up the plot by removing those values:

```{r}
covid_data %>%
  filter(state %in% c("IA","NE","SD","WI","MN","IL","MO")) %>% 
  filter(new_deaths>=0) %>% 
  ggplot(aes(date,new_deaths)) +
  geom_line() +
  facet_wrap(~state, scales = "free_y") +
  ylab("New Daily Deaths") +
  xlab("Date") +
  ggtitle("New daily deaths reported in IA and neighboring states")
```

## Analyze relationship with state population and cases/mortality

Next, create a scatter plot for all 50 states comparing the state population to the total number of casese. Then add a linear trend line using `geom_smooth(method = lm)` , to depicted the "expected" relationship between

```{r}
covid_data %>% 
  filter(date==max(date)) %>% 
  ggplot(aes(x = population, y = total_cases)) +
  geom_point() +
  geom_smooth(method = lm) +
  ylab("Total Cases") +
  xlab("State Population") +
  ggtitle("Trends in population and total cases across all states")
```

Now do the same thing with mortality:

```{r}
covid_data %>% 
  filter(date==max(date)) %>% 
  ggplot(aes(x = population, y = total_deaths)) +
  geom_point() +
  geom_smooth(method = lm) +
  ylab("Total Deaths") +
  xlab("State Population") +
  ylab("Total Deaths") +
  xlab("State Population") +
  ggtitle("Trends in population and total deaths across all states")
```

Next try using a reshaping function to reshape the data and plot both the trends with mortality and number of cases

```{r, message = FALSE}
covid_data %>%
  filter(date==max(date)) %>% 
  select(date,state,population,total_cases,total_deaths) %>% 
  gather(key = key, value = value, -date, - state, -population) %>% 
  mutate(key = ifelse(key=="total_cases","Total Cases", "Total Deaths")) %>% 
  ggplot(aes(x = population, y = value)) +
  geom_point() +
  geom_smooth(method = lm) +
  facet_wrap(~key,scales = "free_y") +
  ylab("Total Cases/Deaths") +
  xlab("State Population") +
  ggtitle("Trends in population and total cases/deaths across all states")
```

**Advanced question (optional):** Can you identify which states have a total number of deaths that fall above or below their expected value based on population? In other words, which states have deaths that fall above the trend line in the plots above. Here is what the resulting values would look like:

```{r}
# fit model
model_data <- covid_data %>% 
  filter(date==max(date))

# fit model
fit <- lm(total_deaths ~ population, data = model_data)

model_data %>% 
  mutate(predicted_deaths=predict(fit)) %>% 
  filter(total_deaths>predicted_deaths) %>% 
  select(state,total_deaths,predicted_deaths) %>% 
  mutate(excess_deaths=total_deaths-predicted_deaths)

```

### How has the pandemic played out in the Midwest compared to other locations in the US across time?

Create a series of plot that describe how the pandemic has played out in the midwest compared to the rest of the US. Specifically compare Iowa and it's neighbors to the rest of the 50 states. To do so create a plot with lines plotting the cumulative number of cases across times for all 50 states. Then make the lines for Iowa and its neighbors colored and thick, while keeping the lines for all other states thin and grey. *Sidenote: These types of plots are often used to depict different simulation trials*

Hint: Here is an example of a way to create a dataset that can then be used to generate this plot:

```{r, echo=TRUE}
plot_data <- covid_data %>% 
  mutate(total_cases_pc = total_cases/population)  %>% 
  select(state,date,total_cases_pc) %>% 
  mutate(state2 = ifelse (state %in% c("IA","NE","SD","WI","MN","IL","MO"), 
                          state, NA),
         total_cases_pc2 = ifelse (state %in% c("IA","NE","SD","WI","MN","IL","MO"),
                                   total_cases_pc, NA)) 

```

Here is what your final plot should look like:

```{r, warning=FALSE}
plot_data %>% 
  ggplot(aes(date,total_cases_pc, group = state)) +
  geom_line(color = "grey", size = 0.3, alpha = 0.7) +
  geom_line(aes(x = date, y=total_cases_pc2, color = state2),size = 1.2) +
  theme_bw() +
  ylab("Total Cases Per Capita") +
  xlab("Reporting Date") +
  theme(legend.title = element_blank()) +
  scale_color_discrete(na.translate=FALSE) +
  ggtitle("Total Cases in Iowa and Neighboring States",
          subtitle = "Other US states depicted in grey")
```

Next, create a similar plot to compare Iowa to a handful of other states across the US. Specifically compare Iowa to AZ, NJ, MA, ND, VT, CO and FL. Your plot should then look like the following:

```{r, warning=FALSE}

plot_data <- covid_data %>% 
  mutate(total_cases_pc = total_cases/population)  %>% 
  select(state,date,total_cases_pc) %>% 
  mutate(state2 = ifelse (state %in% c("IA","AZ","NJ","MA","ND","VT","CO","FL"), state, NA)) %>% 
  mutate(total_cases_pc2 = ifelse (state %in% c("IA","AZ","NJ","MA","ND","VT","CO","FL"), total_cases_pc, NA)) 

plot_data %>% 
  ggplot(aes(date,total_cases_pc, group = state)) +
  geom_line(color = "grey", size = 0.3, alpha = 0.7) +
  geom_line(aes(x = date, y=total_cases_pc2, color = state2),size = 1) +
  theme_bw() +
  ylab("Total Cases Per Capita") +
  xlab("Reporting Date") +
  theme(legend.title = element_blank()) +
  scale_color_discrete(na.translate=FALSE) +
  ggtitle("Total Cases in Iowa and Neighboring States",
          subtitle = "Other US states depicted in grey")
```

**Challenge problem:** Create plots depicting the trajectory of the pandemic across multiple states, like the above plots, but now use facets to create a side-by-side plot of both the number of total cases and total deaths.

Your plots should look like the following for Iowa and neighboring states. (Hint: use `select()` along with a reshaping command before you compute the number of total cases per capita; this will create a column indicator to denote deaths per capita and cases per-capita).

```{r, warning=FALSE}
plot_data <- covid_data %>% 
  select(date,state,total_cases,total_deaths,population) %>% 
  gather(key = key,value = value, -date, -state,-population) %>% 
  mutate(total_cases_pc = value/population)  %>% 
  mutate(state2 = ifelse (state %in% c("IA","NE","SD","WI","MN","IL","MO"), state, NA),
         total_cases_pc2 = ifelse (state %in% c("IA","NE","SD","WI","MN","IL","MO"), total_cases_pc, NA)) %>% 
  mutate(key = ifelse(key=="total_cases","Total Cases", "Total Deaths"))

plot_data %>% 
  ggplot(aes(date,total_cases_pc, group = state)) +
  geom_line(color = "grey", size = 0.3, alpha = 0.7) +
  geom_line(aes(x = date, y=total_cases_pc2, color = state2),size = 1.2) +
  theme_bw() +
  ylab("Totals Per Capita") +
  xlab("Reporting Date") +
  facet_wrap(~key, scales = "free_y") +
  theme(legend.position = "bottom",
        legend.title = element_blank()) +
  scale_color_discrete(na.translate=FALSE)
```

And your plots for Iowa and the other non-neighboring states should look like:

```{r, warning=FALSE}
plot_data <- covid_data %>% 
  select(date,state,total_cases,total_deaths,population) %>% 
  gather(key = key,value = value, -date, -state,-population) %>% 
  mutate(total_cases_pc = value/population)  %>% 
  mutate(state2 = ifelse (state %in% c("IA","AZ","NJ","MA","ND","VT","CO","FL"), state, NA),
         total_cases_pc2 = ifelse (state %in% c("IA","AZ","NJ","MA","ND","VT","CO","FL"), total_cases_pc, NA)) %>% 
  mutate(key = ifelse(key=="total_cases","Total Cases", "Total Deaths"))

plot_data %>% 
  ggplot(aes(date,total_cases_pc, group = state)) +
  geom_line(color = "grey", size = 0.3, alpha = 0.7) +
  geom_line(aes(x = date, y=total_cases_pc2, color = state2),size = 1.2) +
  theme_bw() +
  ylab("Totals Per Capita") +
  xlab("Reporting Date") +
  facet_wrap(~key, scales = "free_y") +
  theme(legend.position = "bottom",
        legend.title = element_blank()) +
  scale_color_discrete(na.translate=FALSE)
```
